\textbf{OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units}

### 2017-02-02 submission/Westcott_OptiClust_mSphere_2016.Rmd
Add tex versions of the manuscript

---
title: "**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
    fig_caption: false
csl: msphere.csl
fontsize: 11pt
geometry: margin=1.0in
---

>>>>

---
title: "**OptiClust: Improved method for assigning amplicon-based sequence data to operational taxonomic units**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
    fig_caption: false
csl: msphere.csl
fontsize: 11pt
geometry: margin=1.0in
---

### 2017-02-02 submission/Westcott_OptiClust_mSphere_2016_orig.tex
Add tex versions of the manuscript

A growing number of algorithms have been developed to cluster sequences
into OTUs. These algorithms can be classified into three general
categories. The first category of algorithms has been termed
closed-reference or phylotyping (13, 14). Sequences are compared to a
reference collection and clustered based on the reference sequences that
they are similar to. This approach is fast; however, the method
struggles when a sequence is similar to multiple reference sequences
that may have different taxonomies and when it is not similar to
sequences in the reference. The second category of algorithms has been
called \emph{de novo} because they assign sequences to OTUs without the
use of a reference (14). These include hierarchical algorithms such as
nearest, furthest, and average neighbor (15) and algorithms that employ
heuristics such as abundance or distance-based greedy clustering as
implemented in USEARCH (16) or VSEARCH (17), Sumaclust, OTUCLUST (18),
and Swarm (19). \emph{De novo} methods are agglomerative and tend to be
more computationally intense. It has proven difficult to know which
method generates the best assignments. A third category of algorithm is
open reference clustering, which is a hybrid approach (3, 14). Here
sequences are assigned to OTUs using closed-reference clustering and
sequences that are not within a threshold of a reference sequence are
then clustered using a \emph{de novo} approach. This category blends the
strengths and weaknesses of the other method and adds the complication
that closed-reference and \emph{de novo} clustering use different OTU
definitions. These algorithms take different approaches to handling
large datasets to minimize the time and memory requirements while
attempting to assign sequences to meaningful OTUs.

Several metrics have emerged for assessing the quality of OTU assignment
algorithms. These have included the time and memory required to run the
algorithm (3, 19--21), agreement between OTU assignments and the
sequences' taxonomy (19, 21--31), sensitivity of an algorithm to
stochastic processes (32), the number of OTUs generated by the algorithm
(22, 33), and the ability to regenerate the assignments made by other
algorithms (3, 34). Unfortunately, these methods fail to directly
quantify the quality of the OTU assignments. An algorithm may complete
with minimal time and memory requirements or generate an idealized
number of OTUs, but the composition of the OTUs could be incorrect.
These metrics also tend to be subjective. For instance, a method may
appear to recapitulate the taxonomy of a synthetic community with known
taxonomic structure, but do a poor job when applied to real communities
with poorly defined taxonomic structure or for sequences that are prone
to misclassification. As an alternative, we developed an approach to
objectively benchmark the clustering quality of OTU assignments (13, 35,
36). This approach counts the number of true positives (TP), true
negatives (TN), false positives (FP), and false negatives (FN) based on
the pairwise distances. Sequence pairs that are within the
user-specified threshold and are clustered together represent TPs and
those in different OTUs are FNs. Those sequence pairs that have a
distance larger than the threshold and are not clustered in the same OTU
are TNs and those in the same OTU are FPs. These values can be
synthesized into a single correlation coefficient, the Matthews'
Correlation Coefficient (MCC), which measures the correlation between
observed and predicted classifications and is robust to cases where
there is an uneven distribution across the confusion matrix (37).
Consistently, the average neighbor algorithm was identified as among the
best or the best algorithm. Other hierarchical algorithms such as
furthest and nearest neighbor, which seek to minimize the number of FP
and FN, respectively, fared significantly worse. The distance-based
greedy clustering as implemented in VSEARCH has also performed well. The
computational resources required to complete the average neighbor
algorithm can be significant for large datasets and so there is a need
for an algorithm that efficiently produces consistently high quality OTU
assignments.

@@

\textbf{\emph{OptiClust algorithm.}} The OptiClust algorithm uses the
pairs of sequences that are within a desired threshold of each other
(e.g.~0.03), a list of all sequence names in the dataset, and the metric
that should be used to assess clustering quality. A detailed description
of the algorithm is provided for a toy dataset in the Supplementary
Material. Briefly, the algorithm starts by placing each sequence either
within its own OTU or into a single OTU. The algorithm proceeds by
interrogating each sequence and re-calculating the metric for the cases
where the sequence stays in its current OTU, is moved to each of the
other OTUs, or is moved into a new OTU. The location that results in the
best clustering quality indicates whether the sequence should remain in
its current OTU or be moved to a different or new OTU. Each iteration
consists of interrogating every sequence in the dataset. Although
numerous options are available for optimizing the clusters and for
assessing the quality of the clusters within the mothur-based
implementation of the algorithm (e.g.~sensitivity, specificity,
accuracy, F1 score, etc.), the default metric for optimization and
assessment is MCC because it includes all four parameters from the
confusion matrix (Figure S1). The algorithm continues until the
optimization metric stabilizes or until it reaches a defined stopping
criteria (Figure S2).

@@

\textbf{\emph{OptiClust-generated OTUs are as stable as those from other
algorithms.}} One concern that many have with \emph{de novo} clustering
algorithms is that their output is sensitive to the initial order of the
sequences because each algorithm must break ties where a sequence could
be assigned to multiple OTUs. An additional concern with the OptiClust
algorithm is that it may stabilize at a local optimum. To evaluate these
concerns we compared the results obtained using ten randomizations of
the order that sequences were given to the algorithm. The median the
coefficient of variation across the six datasets for MCC values obtained
from the replicate clusterings using OptiClust was 0.1\% (Figure 1). We
also measured the coefficient of variation for the number of OTUs across
the six datasets for each method. The median coefficient of variation
for the number of OTUs generated using OptiClust was 0.1\%. Confirming
our previous results, all of the methods we tested were stable to
stochastic processes. Of the methods that involved randomization, the
coefficient of variation for MCC values was considerably smaller with
OptiClust than the other methods and the coefficient of variation for
the number of OTUs was comparable to the other methods. The variation
observed in clustering quality suggested that the algorithm does not
appear to converge to a locally optimum MCC value. More importantly, the
random variation does yield output of a similarly high quality.

\textbf{\emph{Time and memory required to complete Optimization-based
clustering scales efficiently.}} Although not as important as the
quality of clustering, the amount of time and memory required to assign
sequences to OTUs is a legitimate concern. We observed that the time
required to complete the OptiClust algorithm (Figure 1C) paralleled the
number of pairwise distances that were smaller than 0.03. To further
evaluate how the speed and memory usage scaled with the number of
sequences in the dataset, we measured the time required and maximum RAM
usage to cluster 20, 40, 60, 80, and 100\% of the unique sequences from
each of the natural datasets using the OptiClust algorithm (Figure 2).
Within each iteration of the algorithm, each sequence is compared to
every other sequence and each comparison requires a recalculation of the
confusion matrix. This would result in a worst case algorithmic
complexity on the order of N\^{}3, where N is the number of unique
sequences. Because the algorithm only needs to keep track of the
sequence pairs that are within the threshold of each other, it is likely
that the implementation of the algorithm is more efficient. To
empirically determine the algorithmic complexity, we fit a power law
function to the data in Figure 2A. We observed power coefficients
between 1.7 and 2.5 for the marine and human datasets, respectively. The
algorithm requires storing a matrix that contains the pairs of sequences
that are close to each other as well as a matrix that indicates which
sequences are clustered together. The memory required to store these
matrices is on the order of N\^{}2, where N is the number of unique
sequences. In fact, when we fit a power law function to the data in
Figure 2B, the power coefficients were 1.9. Using the four natural
community datasets, doubling the number of sequences in a dataset would
increase the time required to cluster the data by 4 to 8-fold and
increase the RAM required by 4-fold. It is possible that future
improvements to the implementation of the algorithm could improve this
performance.

\textbf{\emph{Cluster splitting heuristic generates OTUs that are as
good as non-split approach.}} We previously described a heuristic to
accelerate OTU assignments where sequences were first classified to
taxonomic groups and within each taxon sequences were assigned to OTUs
using the average neighbor clustering algorithm (13). This method is
similar to open reference clustering except that in our approach all
sequences are subjected to \emph{de novo} clustering following
classification whereas in open reference clustering only those sequences
that cannot be classified are subjected to \emph{de novo} clustering.
Our cluster splitting approach accelerated the clustering and reduced
the memory requirements because the number of unique sequences is
effectively reduced by splitting sequences across taxonomic groups.
Furthermore, because sequences in different taxonomic groups are assumed
to belong to different OTUs they are independent, which permits
parallelization and additional reduction in computation time. Reduction
in clustering quality are encountered in this approach if there are
errors in classification or if two sequences within the desired
threshold belong to different taxonomic groups. It is expected that
these errors would increase as the taxonomic level goes from kingdom to
genus. To characterize the clustering quality, we classified each
sequence at each taxonomic level and calculated the MCC values using
OptiClust, average neighbor, and DGC with VSEARCH when splitting at each
taxonomic level (Figure 3). For each method, the MCC values decreased as
the taxonomic resolution increased; however, the decrease in MCC as not
as large as the difference between clustering methods. As the resolution
of the taxonomic levels increased, the clustering quality remained high,
relative to clusters formed from the entire dataset
(i.e.~kingdom-level). The MCC values when splitting the datasets at the
class and genus levels were within 98.0 and 93.0\%, respectively, of the
MCC values obtained from the entire dataset. These decreases in MCC
value resulted in the formation of as many as 4.7 and 22.5\% more OTUs,
respectively, than were observed from the entire dataset. These errors
were due to the generation of additional false negatives due to
splitting similar sequences into different taxonomic groups. For the
datasets included in the current analysis, the use of the cluster
splitting heuristic was probably not worth the loss in clustering
quality. However, as datasets become larger, it may be necessary to use
the heuristic to clustering the data into OTUs.

@@

The cluster optimization procedure is dependent on the metric that is
chosen for optimization. We employed the MCC because it includes the
four values from a confusion matrix. Other algorithms such as the
furthest neighbor and nearest neighbor algorithms minimize the number of
FP and FN, respectively; however, these suffer because the number of FN
and FP are not controlled, respectively (13, 15). Alternatively, one
could optimize based on the sensitivity, specificity, or accuracy, which
are each based on two values from the confusion matrix or they could
optimize based on the F1 score, which is based on three values from the
confusion matrix. Because these metrics do not balance all four
parameters equally, it is likely that one parameter will dominate in the
optimization procedure. For example, optimizing for sensitivity could
lead to a large number of FPs. More FPs increases the number of OTUs
while more FNs collapses OTUs together. It is difficult to know which is
worse since community richness and diversity are linked to the number of
OTUs. In addition, increasing the number of FNs would overstate the
differences between communities while increasing the number of FPs would
overstate their similarity. Therefore, it is important to jointly
minimize the number of FPs and FNs. With this in mind, we decided to
optimize utilizing the MCC. It is possible that other metrics that
balance the four parameters could be developed and employed for
optimization of the clustering.

The OptiClust algorithm is relatively simple. For each sequence it
effectively asks whether the MCC value will increase if the sequence is
moved to a different OTU including creating a new OTU. If the value does
not change, it remains in the current OTU. The algorithm repeats until
the MCC value stabilizes. Assuming that the algorithm is seeded with
each sequence in a separate OTU, it does not appear that the algorithm
converges to a local optimum. Furthermore, execution of the algorithm
with different random number generator seeds produces OTU assignments of
consistently high quality. Future improvements to the implementation of
the algorithm could provide optimization to further improve its speed
and susceptibility to find a local optimum. Users are encouraged to
repeat the OTU assignment several times to confirm that they have found
the best OTU assignments.

@@

\textbf{Table 1. Description of datasets used to evaluate the OptiClust
algorithm and compare its performance to other algorithms.} Each dataset
contains sequences from the V4 region of the 16S rRNA gene. The number
of distances for each dataset are those that were less than or equal to
0.03. The number of OTUs were determined using the OptiClust algorithm.
The even and staggered datasets were generated by extracting the V4
region from full length reference sequences and the datasets from the
natural communities were generated by sequencing the V4 region using a
Illumina MiSeq with either paired 150 or 250 nt reads.

\begin{longtable}[c]{@{}lcccccc@{}}
\toprule
\textbf{Dataset (Ref.)} & \textbf{Read Length} & \textbf{Samples} &
\textbf{Total Seqs.} & \textbf{Unique Seqs.} & \textbf{Distances} &
\textbf{OTUs}\tabularnewline
\midrule
\endhead
Soil (41) & 150 & 18 & 948,243 & 143,677 & 11,775,167 &
40,216\tabularnewline
Marine (42) & 250 & 7 & 1,384,988 & 75,923 & 12,908,857 &
25,787\tabularnewline
Mice (40) & 250 & 360 & 2,825,495 & 32,447 & 6,988,306 &
2,658\tabularnewline
Human (39) & 250 & 489 & 20,951,841 & 121,281 & 38,544,315 &
11,648\tabularnewline
Even (33, 35) & NA & NA & 1,155,800 & 11,558 & 29,694 &
7,651\tabularnewline
Staggered (33, 35) & NA & NA & 1,156,550 & 11,558 & 29,694 &
7,653\tabularnewline
\bottomrule
\end{longtable}

@@

\includegraphics[width=6.0in]{../results/figures/performance_other.png}

\textbf{Figure S1. The OptiClust algorithm is able to effectively
cluster sequences into OTUs by minimizing or maximizing numerous
metrics.} Plot of MCC (A), number of OTUs (B), and execution times (C)
for the comparison of output from the OptiClust algorithm when to
minimizing or maximizing a variety of parameters when applied to four
natural and two synthetic datasets. Within mothur, OTU assignments can
also be made using other metrics including minimizing false positives
and maximizing the specificity, positive predictive value, and true
negatives; however, these all resulted in sequences being assigned to
separate OTUs, which resulted in no false positives and the maximum
number of true negatives. The error bars indicate the range of values
observed for 10 replicates.

\includegraphics[width=3.5in]{../results/figures/optimization.png}

\textbf{Figure S2. The OptiClust algorithm rapidly converges to optimize
the Matthew's Correlation Coefficient.} The six datasets were clustered
into OTUs using the OptiClust algorithm seeking to maximize the
Matthew's Correlation Coefficient. This was repeated 10 times for each
dataset.

>>>>

A growing number of algorithms have been developed to cluster sequences
into OTUs. These algorithms can be classified into three general
categories. The first category of algorithms has been termed
closed-reference or phylotyping (13, 14). Sequences are compared to a
reference collection and clustered based on the reference sequences that
they are similar to. This approach is fast; however, the method
struggles when a sequence is similar to multiple reference sequences
that may have different taxonomies and when it is not similar to
sequences in the reference. The second category of algorithms has been
called \emph{de novo} because they assign sequences to OTUs without the
use of a reference (14). These include hierarchical algorithms such as
nearest, furthest, and average neighbor (15) and algorithms that employ
heuristics such as abundance or distance-based greedy clustering as
implemented in USEARCH (16) or VSEARCH (17), Sumaclust, OTUCLUST (18),
and Swarm (19). \emph{De novo} methods tend to be more computationally
intense and it has proven difficult to know which method generates the
best assignments. A third category of algorithm is open-reference
clustering, which is a hybrid approach (3, 14). Here sequences are
assigned to OTUs using closed-reference clustering and sequences that
are not within a threshold of a reference sequence are then clustered
using a \emph{de novo} approach. This category blends the strengths and
weaknesses of the other method and adds the complication that
closed-reference and \emph{de novo} clustering use different OTU
definitions. These algorithms take different approaches to handling
large datasets to minimize the time and memory requirements while
attempting to assign sequences to meaningful OTUs.

Several metrics have emerged for assessing the quality of OTU assignment
algorithms. These have included the time and memory required to run the
algorithm (3, 19--21), agreement between OTU assignments and the
sequences' taxonomy (19, 21--31), sensitivity of an algorithm to
stochastic processes (32), the number of OTUs generated by the algorithm
(22, 33), and the ability to regenerate the assignments made by other
algorithms (3, 34). Unfortunately, these methods fail to directly
quantify the quality of the OTU assignments. An algorithm may complete
with minimal time and memory requirements or generate an idealized
number of OTUs, but the composition of the OTUs could be incorrect.
These metrics also tend to be subjective. For instance, a method may
appear to be recapitulate the taxonomy of a synthetic community with
known taxonomic structure, but do a poor job when applied to real
communities with poorly defined taxonomic structure or for sequences
that are prone to misclassification. As an alternative, we developed an
approach to objectively benchmark the clustering quality of OTU
assignments (13, 35, 36). This approach counts the number of true
positives (TP), true negatives (TN), false positives (FP), and false
negatives (FN) based on the pairwise distances. Sequence pairs that are
within the user-specified threshold and are clustered together represent
TPs and those in different OTUs are FNs. Those sequence pairs that have
a distance larger than the threshold and are not clustered in the same
OTU are TNs and those in the same OTU are FPs. These values can be
synthesized into a single correlation coefficient, the Matthews'
Correlation Coefficient (MCC), which measures the correlation between
observed and predicted classifications and is robust to cases where
there is an uneven distribution across the confusion matrix (37).
Consistently, the average neighbor algorithm was identified as among the
best or the best algorithm. The distance-based greedy clustering as
implemented in VSEARCH has also performed well. The computational
resources required to complete the average neighbor algorithm can be
significant for large datasets and so there is a need for an algorithm
that efficiently produces consistently high quality OTU assignments.

@@

\textbf{\emph{OptiClust algorithm.}} The OptiClust algorithm uses the
pairs of sequences that are within a desired threshold of each other
(e.g.~0.03), a list of all sequence names in the dataset, and the metric
that should be used to assess clustering quality. A detailed description
of the algorithm is provided for a toy dataset in the Supplementary
Material. Briefly, the algorithm starts by placing each sequence either
within its own OTU or into a single OTU. The algorithm proceeds by
interrogating each sequence and re-calculating the metric for the cases
where the sequence stays in its current OTU, is moved to each of the
other OTUs, or is moved into a new OTU. The location that results in the
best clustering quality indicates whether the sequence should remain in
its current OTU or be moved to a different or new OTU. Each iteration
consists of interrogating every sequence in the dataset. Although
numerous options are available within the mothur-based implementation of
the algorithm (e.g.~sensitivity, specificity, accuracy, F1 score, etc.),
the default metric is MCC because it includes all four parameters from
the confusion matrix. The algorithm continues until the optimization
metric stabilizes or until it reaches a defined stopping criteria.

@@

\textbf{\emph{OptiClust-generated OTUs are as stable as those from other
algorithms.}} One concern that many have with \emph{de novo} clustering
algorithms is that their output is sensitive to the initial order of the
sequences. An additional concern with the OptiClust algorithm is that it
may stabilize at a local optimum. To evaluate these concerns we compared
the results obtained using ten randomizations of the order that
sequences were given to the algorithm. The median the coefficient of
variation across the six datasets for MCC values obtained from the
replicate clusterings using OptiClust was 0.1\% (Figure 1). We also
measured the coefficient of variation for the number of OTUs across the
six datasets for each method. The median coefficient of variation for
the number of OTUs generated using OptiClust was 0.1\%. Confirming our
previous results, all of the methods we tested were stable to stochastic
processes. Of the methods that involved randomization, the coefficient
of variation for MCC values considerably smaller with OptiClust than the
other methods and the coefficient of variation for the number of OTUs
was comparable to the other methods. The variation observed in
clustering quality suggested that the algorithm does not appear to
converge to a locally optimum MCC value. More importantly, the random
variation does yield output of a similarly high quality.

\textbf{\emph{Time and memory required to complete Optimization-based
clustering scales efficiently.}} Although not as important as the
quality of clustering, the amount of time and memory required to assign
sequences to OTUs is a legitimate concern. To evaluate how the speed and
memory usage scaled with the number of sequences in the dataset, we
measured the time required and maximum RAM usage to cluster 20, 40, 60,
80, and 100\% of the unique sequences from each of the natural datasets
using the OptiClust algorithm (Figure 2). Within each iteration of the
algorithm, each sequence is compared to every other sequence and each
comparison requires a recalculation of the confusion matrix. This would
result in a worst case algorithmic complexity on the order of N\^{}3,
where N is the number of unique sequences. Because the algorithm only
needs to keep track of the sequence pairs that are within the threshold
of each other, it is likely that the implementation of the algorithm is
more efficient. To empirically determine the algorithmic complexity, we
fit a power law function to the data in Figure 2A. We observed power
coefficients between 1.7 and 2.5 for the marine and human datasets,
respectively. The algorithm requires storing a matrix that contains the
pairs of sequences that are close to each other as well as a matrix that
indicates which sequences are clustered together. The memory required to
store these matrices is on the order of N\^{}2, where N is the number of
unique sequences. In fact, when we fit a power law function to the data
in Figure 2B, the power coefficients were 1.9. This analysis suggests
that doubling the number of sequences in a dataset would increase the
time required to cluster the data by 4 to 8-fold and increase the RAM
required by 4-fold. It is possible that future improvements to the
implementation of the algorithm could improve this performance.

\textbf{\emph{Cluster splitting heuristic generates OTUs that are as
good as non-split approach.}} We previously described a heuristic to
accelerate OTU assignments where sequences were first classified to
taxonomic groups and within each taxon sequences were assigned to OTUs
using the average neighbor clustering algorithm (13). This accelerated
the clustering and reduce the memory requirements because the number of
unique sequences is effectively reduced by splitting sequences across
taxonomic groups. Furthermore, because sequences in different taxonomic
groups are assumed to belong to different OTUs they are independent,
which permits parallelization and additional reduction in computation
time. Reduction in clustering quality are encountered in this approach
if there are errors in classification or if two sequences within the
desired threshold belong to different taxonomic groups. It is expected
that these errors would increase as the taxonomic level goes from
kingdom to genus. To characterize the clustering quality, we calculated
the MCC values using OptiClust, average neighbor, and DGC with VSEARCH
when splitting at each taxonomic level (Figure 3). For each method, the
MCC values decreased as the taxonomic resolution increased; however, the
decrease in MCC as not as large as the difference between clustering
methods. As the resolution of the taxonomic levels increased, the
clustering quality remained high, relative to clusters formed from the
entire dataset (i.e.~kingdom-level). The MCC values when splitting the
datasets at the class and genus levels were within 98.0 and 93.0\%,
respectively, of the MCC values obtained from the entire dataset. These
decreases in MCC value resulted in the formation of as many as 4.7 and
22.5\% more OTUs, respectively, than were observed from the entire
dataset. For the datasets included in the current analysis, the use of
the cluster splitting heuristic was probably not worth the loss in
clustering quality. However, as datasets become larger, it may be
necessary to use the heuristic to clustering the data into OTUs.

@@

The cluster optimization procedure is dependent on the metric that is
chosen for optimization. We employed the MCC because it includes the
four values from a confusion matrix. Other algorithms such as the
furthest neighbor and nearest neighbor algorithms minimize the number of
FP and FN, respectively; however, these suffer because the number of FN
and FP are not controlled (13, 15). Alternatively, one could optimize
based on the sensitivity, specificity, or accuracy, which are each based
on two values from the confusion matrix or they could optimize based on
the F1 score, which is based on three values from the confusion matrix.
Because these metrics do not balance all four parameters equally, it is
likely that one parameter will dominate in the optimization procedure.
For example, optimizing for sensitivity could lead to a large number of
FPs. Since we would like to minimize both FPs and FNs and not just the
total number of false assignments, we decided to optimize utilizing the
MCC. It is possible that other metrics could be developed and employed
for optimization of the clustering.

The OptiClust algorithm is relatively simple. For each sequence it
effectively ask whether the MCC value will increase if the sequence is
moved to a different OTU including creating a new OTU. If the value does
not change, it remains in the current OTU. The algorithm repeats until
the MCC value stabilizes. Assuming that the algorithm is seeded with
each sequence in a separate OTU, it does not appear that the algorithm
converges to a local optimum. Furthermore, execution of the algorithm
with different random number generator seeds produces OTU assignments of
consistently high quality. Future improvements to the implementation of
the algorithm could provide optimization to further improve its speed
and susceptibility to find a local optimum. Users are encourage to
repeat the OTU assignment several times to confirm that they have found
the best OTU assignments.

@@

\textbf{Table 1. Description of datasets used to evaluate the OptiClust
algorithm and compare its performance to other algorithms.} Each dataset
contains sequences from the V4 region of the 16S rRNA gene. The even and
staggered datasets were generated by extracting the V4 region from full
length reference sequences and the datasets from the natural communities
were generated by sequencing the V4 region using a Illumina MiSeq with
either paired 150 or 250 nt reads.

\begin{longtable}[c]{@{}lcccc@{}}
\toprule
\textbf{Dataset (Ref.)} & \textbf{Read Length} & \textbf{Samples} &
\textbf{Total Seqs.} & \textbf{Unique Seqs.}\tabularnewline
\midrule
\endhead
Soil (41) & 150 & 18 & 948,243 & 143,677\tabularnewline
Marine (42) & 250 & 7 & 1,384,988 & 75,923\tabularnewline
Mice (40) & 250 & 360 & 2,825,495 & 32,447\tabularnewline
Human (39) & 250 & 489 & 20,951,841 & 121,281\tabularnewline
Even (33, 35) & NA & NA & 1,155,800 & 11,558\tabularnewline
Staggered (33, 35) & NA & NA & 1,156,550 & 11,558\tabularnewline
\bottomrule
\end{longtable}

@@
